\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2020-08-05 analysis/paper/paper.tex
update cs 3&4

\author{
Sahir Rai Bhatnagar*\\McGill University \And Maxime Turgeon*\\University of Manitoba \AND Jesse Islam\\McGill University \And James Hanley\\McGill University \And Olli Saarela\\University of Toronto
}
\title{\pkg{casebase}: An Alternative Framework For Survival Analysis}

\Plainauthor{Sahir Rai Bhatnagar*, Maxime Turgeon*, Jesse Islam, James Hanley, Olli Saarela}
\Plaintitle{casebase: An Alternative Framework For Survival Analysis}
\Shorttitle{\pkg{casebase}: An Alternative Framework For Survival Analysis}

@@

      James Hanley\\
  McGill University\\
  1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
  E-mail: \email{james.hanley@mcgill.ca}\\
  URL: \url{http://www.medicine.mcgill.ca/epidemiology/hanley/}\\~\\
      Olli Saarela\\
  University of Toronto\\
  Dalla Lana School of Public Health, 155 College Street, 6th floor,
  Toronto, Ontario M5T 3M7, Canada\\
  E-mail: \email{olli.saarela@utoronto.ca}\\
  URL: \url{http://individual.utoronto.ca/osaarela/}\\~\\
  }

@@

Survival analysis has been greatly influenced over the last 50 years by
the partial likelihood approach of the Cox proportional hazard model
\citep{cox1972regression}. This approach provides a flexible way of
assessing the influence of covariates on the hazard function, without
the need to specify a parametric survival model. This flexibility comes
at the cost of decoupling the baseline hazard from the effect of the
covariates. To recover the whole survival curve---or the cumulative
incidence function (CIF)---we then need to separately estimate the
baseline hazard \citep{breslow1972discussion}. This in turn often leads
to stepwise estimates of the survival function.

@@

With the goal of fitting smooth-in-time hazard functions, Hanley \&
Miettinen \citeyearpar{hanley2009fitting} proposed a general framework
for estimating fully parametric hazard models via logistic regression.
Their approach provides users that are familiar with generalized linear
models with a natural way of fitting parametric survival models.
Moreover, their framework is very flexible: general functions of time
can be estimated (e.g.~using splines or general additive models), and
hence these models retain some of the flexibility of partial likelihood
approaches.

In this article, we present an \proglang{R} package that combines the
ideas of Hanley \& Miettinen into a simple interface. The purpose of the
\pkg{casebase} package is to provide practitioners with an easy-to-use
software tool to predict the risk (or cumulative incidence) of an event,
conditional on a particular patient's covariate profile. Our package
retains the flexibility of case-base sampling and the familiar interface
of the \code{glm} function. In addition, we provide extensive
visualization tools.

@@

As discussed in Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the
key idea behind case-base sampling is to discretize the study base into
an infinite amount of \emph{person moments}. These person moments are
indexed by both an individual in the study and a time point, and
therefore each person moment has a covariate profile, an exposure status
and an outcome status attached to it. We note that there is only a
finite number of person moments associated with the event of interest
(what Hanley \& Miettinen call the \emph{case series}). Case-base
sampling refers to the sampling from the base of a representative finite
sample called the \emph{base series}.

To describe the theoretical foundations of case-base sampling, we use
the framework of counting processes. In what follows, we abuse notation
slightly and omit any mention of \(\sigma\)-algebras; the interested
reader can refer to Saarela \& Arjas \citeyearpar{saarela2015non} and
Saarela \citeyearpar{saarela2016case}. First, let
\(N_{i}(t) \in \{0, 1\}\) be counting processes corresponding to the
event of interest for individual \(i=1, \ldots,n\). For simplicity, we
will consider Type I censoring due to the end of follow-up at time
\(\tau\) (the general case of non-informative censoring is treated in
Saarela \citeyearpar{saarela2016case}). We assume a continuous time
model, which implies that the counting process jumps are less than or
equal to one. We are interested in modeling the hazard functions
\(\lambda_{i}(t)\) of the processes \(N_i(t)\), and which satisfy
\[\lambda_{i}(t) dt = E[dN_{i}(t)].\]

Next, we model the base series sampling mechanism using non-homogeneous
Poisson processes \(R_i(t) \in \{0, 1, 2, \ldots\}\), with the
person-moments where \(dR_i(t) = 1\) constituting the base series. The
process \(Q_{i}(t) = R_i(t) + N_{i}(t)\) then counts both the case and
base series person-moments contributed by individual \(i\). This process
is typically defined by the user via its hazard function \(\rho_i(t)\).
The process \(Q_{i}(t)\) is characterized by
\(E[dQ_{i}(t)] = \lambda_{i}(t)dt + \rho_i(t)dt\).

If the hazard function \(\lambda_{i}(t; \theta)\) is parametrized in
terms of \(\theta\), we could define an estimator \(\hat{\theta}\) by
maximization of the likelihood expression
\[L_0(\theta) = \prod_{i=1}^n \exp\left\{ -\int_0^\tau \lambda_i(t; \theta) dt \right\} \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \lambda_{i}(t;\theta)^{dN_{i}(t)}\left(1 - \lambda_{i}(t;\theta)\right)^{1 - dN_{i}(t)},\]
where \(\prod_{t\in[0,u)}\) represents a product integral from \(0\) to
\(u\). However, the integral over time makes the computation and
maximization of \(L_0(\theta)\) challenging.

Case-base sampling allows us to avoid this integral. By conditioning on
a sampled person-moment, we get individual contributions of the form
\[P(dN_{i}(t) \mid dQ_{i}(t) = 1) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.\]
Therefore, we can define an estimating equation for \(\theta\) as
follows:
\[L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{dQ_i(t)}.\]
When a logarithmic link function is used for modeling the hazard
function, the above expression is of a logistic regression form with an
offset term \(\log(1/\rho_i(t))\). Note that the sampling units selected
in the case-base sampling mechanism are person-moments, rather than
individuals, and the parameters to be estimated are hazards or hazard
ratios rather than odds or odds ratios. Generally, an individual can
contribute more than one person-moment, and thus the terms in the
product integral are not independent. Nonetheless, Saarela
\citeyearpar{saarela2016case} showed that the logarithm of this
estimating function has mean zero at the true value \(\theta=\theta_0\),
and that the resulting estimator \(\hat{\theta}\) is asymptotically
normally distributed.

@@

Case-base sampling can also be used in the context of competing-risk
analyses. Assuming there are \(J\) competing events, we can show that
each person-moment's contribution of the likelihood is of the form

@@

where \(N_j(t)\) is the counting process associated with the event of
type \(j\) and \(\lambda_j(t)\) is the corresponding hazard function. As
may be expected, this functional form is similar to the terms appearing
in the likelihood function for multinomial
regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1.\end{align*}}

@@

Parametric survival models are implemented in a handful of packages:
\pkg{CFC} \citeyearpar{mahani2015bayesian}, \pkg{flexsurv}
\citeyearpar{flexsurv}, \pkg{SmoothHazard} \citeyearpar{smoothHazard},
\pkg{rsptm2} \citeyearpar{clements_liu}, \pkg{mets}
\citeyearpar{scheike2014estimating}, and \pkg{survival}
\citeyearpar{survival-package}. The types of models they allow vary for
each package. For example, \pkg{SmoothHazard} is limited to Weibull
distributions \citeyearpar{smoothHazard}, whereas both \pkg{flexsurv}
and \pkg{survival} allow users to supply any distribution of their
choice. Also, \pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets} and
\pkg{rstpm2} also have the ability to model the effect of time using
splines, which allows flexible modeling of the hazard function.
Moreover, \pkg{flexsurv} has the ability to estimate both scale and
shape parameters for a variety of parametric families. As discussed
above, \pkg{casebase} can model any parametric family whose log-hazard
can be expressed as a linear combination of covariates (including time).
Therefore, our package allows the user to model the effect of time using
splines. Also, by including interaction terms between covariates and
time, it also allows users to fit time-varying coefficient models.
However, we do not explicitly model any shape parameter, unlike
\pkg{flexsurv}.

Several packages implement penalized estimation for the Cox model:
\pkg{glmnet} \citeyearpar{regpathcox}, \pkg{glmpath}
\citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
\pkg{RiskRegression} \citeyearpar{gerds_blanche}. Moreover, some
packages also include penalized estimation in the context of Cox models
with time-varying coefficients: elastic-net penalization with
\pkg{CoxRidge} \citeyearpar{perperoglou} and \pkg{rstpm2}
\citeyearpar{clements_liu}, while \pkg{survival}
\citeyearpar{survival-package} has an implementation of ridge-penalized
estimation. On the other hand, our package \pkg{casebase} provides
penalized estimation in the context of parametric hazards. To our
knowledge, \pkg{casebase} and \pkg{rsptm2} are the only packages to
offer this functionality.

Next, several \proglang{R} packages implement methodologies for
competing risk analysis; for a different perspective on this topic, see
Mahani \& Sharabiani \citeyearpar{mahani2015bayesian}. The package
\pkg{cmprsk} provides methods for cause-specific subdistributions, such
as in the Fine-Gray model \citeyearpar{fine1999proportional}. On the
other hand, the package \pkg{CFC} estimates cause-specific CIFs from
unadjusted, non-parametric survival functions. Our package
\pkg{casebase} also provides functionalities for competing risk analysis
by estimating parametrically the cause-specific hazards. From these
quantities, we can then estimate the cause-specific CIFs.

Finally, several packages include functions to estimate the CIF. The
corresponding methods generally fall into two categories: transformation
of the estimated hazard function, and semi-parametric estimation of the
baseline hazard. The first category broadly corresponds to parametric
survival models, where the full hazard is explicitly modeled. Using this
estimate, the survival function and the CIF can be obtained using their
functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI}
below). Packages providing this functionality include \pkg{CFC},
\pkg{Flexsurv}, \pkg{mets}, and \pkg{survival}. Our package
\pkg{casebase} also follows this approach for both single-event and
competing-risk analyses. The second category outlined above broadly
corresponds to semi-parametric models. These models do not model the
full hazard function, and therefore the baseline hazard needs to be
estimated separately in order to estimate the survival function. This is
achieved using semi-parametric estimators (e.g.~Breslow's estimator) or
parametric estimators (e.g.~spline functions). Packages that implement
this approach include \pkg{RiskRegression}, \pkg{rstpm2}, and
\pkg{survival}. As mentioned in the introduction, a key distinguishing
factor between these two approaches is that the first category leads to
smooth estimates of the cumulative incidence function, whereas the
second category often produces estimates in the form of step-wise
functions. Providing smooth estimates of the CIF was one of the main
motivations for introducing case-base sampling in survival analysis.

@@

Figure \ref{fig:plot-stratified-erspc-data} is built sequentially by
first adding a layer for the area representing the population time in
gray, with subjects having the least amount of observation time plotted
at the top of the y-axis. We immediately notice a distinctive
\emph{step-wise shape} in the population time area. This is due to the
randomization of the Finnish cohorts which were carried out on January 1
of each of year from 1996 to 1999. Coupled with the uniform December 31
2006 censoring date, this led to large numbers of men with exactly 11,
10, 9 or 8 years of follow-up. Tracked backwards in time (i.e.~from
right to left), the population-time plot shows the recruitment pattern
from its beginning in 1991, and the January 1 entries in successive
years. Tracked forwards in time (i.e.~from left to right), the plot for
the first three years shows attrition due entirely to death (mainly from
other causes). Since the Swedish and Belgian centres were the last to
complete recruitment--in December 2003--the minimum potential follow-up
is three years. Tracked further forwards in time (i.e.~after year 3) the
attrition is a combination of deaths and staggered entries. As we can
see, population-time plots summarise a wealth of information about the
study into a simple graph.

@@

\[ S(t) = \exp\left(-\int_0^t \Lambda(u;X) du\right),\qquad \Lambda(t;X) = \sum_{j=1}^J \lambda_j(t;X).\]
From this, we can derive the event-specific subdensities:

@@

The data contains information on 177 patients who received a stem-cell
transplant for acute leukemia. The event of interest is relapse, but
other competing causes (e.g.~transplant-related death) were also
recorded. Several covariates were captured at baseline: sex, disease
type (acute lymphoblastic or myeloblastic leukemia, abbreviated as ALL
and AML, respectively), disease phase at transplant (Relapse, CR1, CR2,
CR3), source of stem cells (bone marrow and peripheral blood, coded as
BM+PB, or only peripheral blood, coded as PB), and age.

@@

We will compare our estimates to that obtained from a corresponding
Fine-Gray model \citeyearpar{fine1999proportional}. The Fine-Gray model
is a semiparametric model for the cause-specific \emph{subdistribution},
i.e.~the function \(f_k(t)\) such that
\[CI_k(t) =1 - \exp\left( - \int_0^t f_k(u) \textrm{d}u \right),\] where
\(CI_k(t)\) is the cause-specific cumulative incidence. The Fine-Gray
model allows to directly assess the effect of a covariate on the
subdistribution, as opposed to the hazard. For the computation, we will
use the \pkg{timereg} package \citep{timereg}:

@@

\caption{\label{fig:brier1} Comparison of three Cox regression models to the unadjusted Kaplan-Meier curve. (A) Cumulative Incidence curves. For a given model, the plotted curve is the average curve of each individual in the study. Each follow-up time seen in the study was used. There are minimal differences between the curves. APS coincides with the SPS curve, as they are considerably overlapped.(B) Time-dependent Brier scores, where a lower score is better. The full model performs best overall.}\label{fig:brierpt1}
\end{figure}
\end{CodeChunk}

@@

\caption{\label{fig:cs3lolliPlot} Lollipop plot showing the coefficient estimates between three models. If a coefficient estimate is 0 in any model, it is removed from the plot for simplicity. Shrinkage occurs due to penalization.}\label{fig:coefplots}
\end{figure}
\end{CodeChunk}

These values do not inform us of model performance. To do so, we take
our hazard functions and use them to predict the absolute risk of
everyone in our test set, over time. The probabilities over time for
each individual are averaged, resulting in the absolute risk curves in
Figure \ref{fig:cs3FinalBrier} (A). There is separation between the
curves, but it is minimal. To better demonstrate the differences in
performance between the models, we compare Brier scores over time. In
Figure \ref{fig:cs3FinalBrier} (B), the Brier score is lower in our
penalized logistic regression model compared to both Cox regression
models. The differences, however, are minimal. All three outperform the
Kaplan-Meier model.

@@

\caption{\label{fig:cs3FinalBrier} (A) Four curves comparing the probabilities of death as it changes over follow-up time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier. All follow-up times seen in the new data were used. (B) Four curves demonstrating the Brier scores as they change over time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier. Lower is better. Penalized logistic regression produced the lowest results over follow-up time.}\label{fig:riskregressionBrier}
\end{figure}
\end{CodeChunk}

@@

As before, we can look at the population-time plot for a graphical
summary of the event incidence. As we can see, most events occur early
during the follow-up period, and therefore we do not expect the hazard
to be constant.

@@

Note that the fourth model includes an interaction term between exposure
and follow-up time. In other words, this model no longer exhibit
proportional hazards. The evidence of non-proportionality of hazards in
the Stanford Heart Transplant data has been widely discussed
\citep{arjas1988graphical}.

@@

\begin{CodeOutput}
#> Model1 Model2 Model3 Model4 
#>    491    459    444    445
\end{CodeOutput}
\end{CodeChunk}

As we can, the best fit is the fourth model. By visualizing the hazard
functions for both exposed and unexposed individuals, we can more
clearly see how the hazards are no longer proportional.

@@

\begin{CodeInput}
R> # Compute hazards---
R> # First, create a list of time points for both exposure status
R> hazard_data <- expand.grid(
R+   exposure = c(0, 1),
R+   futime = seq(0, 425,
R+     length.out = 100
R+   )
R+ )
R> 
R> # Set the offset to zero
R> hazard_data$offset <- 0
R> # Use predict to get the fitted values, and exponentiate to
R> # transform to the right scale
R> hazard_data$hazard <- exp(predict(fit4,
R+   newdata = hazard_data,
R+   type = "link"
R+ ))
R> # Add labels for plots
R> hazard_data$Status <- factor(hazard_data$exposure,
R+   labels = c("NoTrans", "Trans")
R+ )
R> 
R> ggplot(hazard_data, aes(futime, hazard, colour = Status)) +
R+   geom_line() +
R+   paper_gg_theme +  
R+   ylab("Hazard") +
R+   xlab("Follow-up time")
\end{CodeInput}

@@

Note that we can easily adapt the code above to the situation where a
patient receives a heart transplant at a point in time of interest, for
example after 30 days.

\begin{center}\includegraphics{../figures/stanford-1y-haz-1} \end{center}

We can then compare the 1-year mortality risk without transplant and
with transplant at 30 days.

\begin{CodeOutput}
#>  time     
#>     0 0.00
#>   365 0.72
\end{CodeOutput}

\begin{CodeOutput}
#> [1] 0.65
\end{CodeOutput}
\end{CodeChunk}

As we can see, the risk estimate at 1-year is about 30\% lower if the
patient receives a heart transplant at 30 days.

@@

\begin{center}\includegraphics{../figures/unnamed-chunk-12-1} \end{center}

@@

@@

This report was generated on 2020-07-28 09:55:47 using the following
computational environment and dependencies:

>>>>

\author{
Sahir Rai Bhatnagar*\\McGill University \And Maxime Turgeon*\\University
of Manitoba \AND Jesse Islam\\McGill University \And James A.
Hanley\\McGill University \And Olli Saarela\\University of Toronto
}
\title{\pkg{casebase}: An Alternative Framework For Survival Analysis
and Comparison of Event Rates}

\Plainauthor{Sahir Rai Bhatnagar*, Maxime Turgeon*, Jesse Islam, James
A. Hanley, Olli Saarela}
\Plaintitle{casebase: An Alternative Framework For Survival Analysis and
Comparison of Event Rates}
\Shorttitle{\pkg{casebase}: An Alternative Framework For Survival
Analysis}

@@

      James A. Hanley\\
  McGill University\\
  1020 Pine Avenue West Montreal, QC, Canada H3A 1A2\\
  E-mail: \email{james.hanley@mcgill.ca}\\
  URL: \url{http://www.medicine.mcgill.ca/epidemiology/hanley/}\\~\\
      Olli Saarela\\
  University of Toronto\\
  Dalla Lana School of Public Health, 155 College Street, 6th floor,
Toronto, Ontario M5T 3M7, Canada\\
  E-mail: \email{olli.saarela@utoronto.ca}\\
  URL: \url{http://individual.utoronto.ca/osaarela/}\\~\\
  }

@@

Survival analysis and the comparison of event rates has been greatly
influenced over the last 50 years by the partial likelihood approach of
the Cox proportional hazard model \citep{cox1972regression}. This
approach provides a flexible way of assessing the influence of
covariates on the hazard function, without the need to specify a
parametric survival model. This flexibility comes at the cost of
decoupling the baseline hazard from the effect of the covariates. To
recover the whole survival curve---or the cumulative incidence function
(CIF)---we then need to separately estimate the baseline hazard
\citep{breslow1972discussion}. This in turn leads to stepwise estimates
of the survival function.

@@

With the goal of fitting smooth-in-time hazard functions, Hanley \&
Miettinen \citeyearpar{hanley2009fitting} proposed a general framework
for estimating fully parametric hazard models via logistic regression.
Their approach provides users that are familiar with generalized linear
models with a natural way of fitting parametric survival models.
Moreover, their framework is very flexible: general functions of time
can be estimated (e.g.~using splines or general additive models), and
hence these models retain some of the flexibility of Cox's partial
likelihood approach.

In this article, we present an \proglang{R} package that combines the
ideas of Hanley \& Miettinen into a simple interface. The purpose of the
\pkg{casebase} package is to provide practitioners with an easy-to-use
software tool to compute a patient's risk (or cumulative incidence) of
an event, conditional on a particular patient's covariate profile. Our
package retains the flexibility of case-base sampling and the familiar
interface of the \code{glm} function. In addition, we provide extensive
visualization tools.

@@

As discussed in Hanley \& Miettinen \citeyearpar{hanley2009fitting}, the
key idea behind case-base sampling is to sample from the study base a
finite amount of \emph{person moments}. These person moments are indexed
by both an individual in the study and a time point, and therefore each
person moment has a covariate profile, an exposure status and an outcome
status attached to it. We note that there is only a finite number of
person moments associated with the event of interest (what Hanley \&
Miettinen call the \emph{case series}). Case-base sampling refers to the
sampling from the base of a representative finite sample called the
\emph{base series}.

To describe the theoretical foundations of case-base sampling, we use
the framework of counting processes. In what follows, we abuse notation
slightly and omit any mention of \(\sigma\)-algebras. Instead, following
Aalen \emph{et al} \citeyearpar{aalen2008survival}, we use the
placeholder \(\mathrm{past}\) to denote the past history of the
corresponding process. The reader interested in more details can refer
to Saarela \& Arjas \citeyearpar{saarela2015non} and Saarela
\citeyearpar{saarela2016case}. First, let \(N_{i}(t) \in \{0, 1\}\) be
counting processes corresponding to the event of interest for individual
\(i=1, \ldots,n\). For simplicity, we will consider Type I censoring due
to the end of follow-up at time \(\tau\) (the general case of
non-informative censoring is treated in Saarela
\citeyearpar{saarela2016case}). We assume a continuous time model, which
implies that the counting process jumps are less than or equal to one.
We are interested in modeling the hazard functions \(\lambda_{i}(t)\) of
the processes \(N_i(t)\), and which satisfy
\[\lambda_{i}(t) dt = E[dN_{i}(t)\mid \mathrm{past}].\]

Next, we model the base series sampling mechanism using non-homogeneous
Poisson processes \(R_i(t) \in \{0, 1, 2, \ldots\}\), with the
person-moments where \(dR_i(t) = 1\) constituting the base series. The
process \(Q_{i}(t) = R_i(t) + N_{i}(t)\) then counts both the case and
base series person-moments contributed by individual \(i\). This process
is typically defined by the user via its intensity function
\(\rho_i(t)\). The process \(Q_{i}(t)\) is characterized by
\(E[dQ_{i}(t)\mid\mathrm{past}] = \lambda_{i}(t)dt + \rho_i(t)dt\).

If the hazard function \(\lambda_{i}(t; \theta)\) is parametrized in
terms of \(\theta\), we could define an estimator \(\hat{\theta}\) by
maximization of the likelihood expression
\[L_0(\theta) = \prod_{i=1}^n \exp\left\{ -\int_0^{\min(t_i,\tau)} \lambda_i(t; \theta) dt \right\} \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \lambda_{i}(t;\theta)^{dN_{i}(t)},\]
where \(\prod_{t\in[0,u)}\) represents a product integral from \(0\) to
\(u\), and where \(t_i\) is the event time for individual \(i\).
However, the integral over time makes the computation and maximization
of \(L_0(\theta)\) challenging.

Case-base sampling allows us to avoid this integral. By conditioning on
a sampled person-moment, we get individual likelihood contributions of
the form
\[P(dN_{i}(t) \mid dQ_{i}(t) = 1,\mathrm{past}) \stackrel{\theta}{\propto} \frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}.\]
Therefore, we can define an estimating function for \(\theta\) as
follows:
\[L(\theta) = \prod_{i=1}^{n} \prod_{t\in[0,\tau)} \left(\frac{\lambda_{i}(t; \theta)^{dN_{i}(t)}}{\rho_i(t) + \lambda_{i}(t;\theta)}\right)^{dQ_i(t)}.\]
When a logarithmic link function is used for modeling the hazard
function, the above expression is of a logistic regression form with an
offset term \(\log(1/\rho_i(t))\). Note that the sampling units selected
in the case-base sampling mechanism are person-moments, rather than
individuals, and the parameters to be estimated are hazards or hazard
ratios rather than odds or odds ratios. Generally, an individual can
contribute more than one person-moment, and thus the terms in the
product integral are not independent. Nonetheless, Saarela
\citeyearpar{saarela2016case} showed that the logarithm of this
estimating function has mean zero at the true value \(\theta=\theta_0\),
and that the resulting estimator \(\hat{\theta}\) is asymptotically
normally distributed.

@@

Case-base sampling can also be used in the context of competing-risk
analyses. Assuming there are \(J\) competing events, we can show that
each person-moment's contribution to the likelihood is of the form

@@

where \(N_j(t)\) is the counting process associated with the event of
type \(j\) and \(\lambda_j(t)\) is the corresponding cause-specific
hazard function. As may be expected, this functional form is similar to
the terms appearing in the likelihood function for multinomial
regression.\footnote{Specifically, it corresponds to the following parametrization: \begin{align*} \log\left(\frac{P(Y=j \mid X)}{P(Y = J \mid X)}\right) = X^T\beta_j, \qquad j = 1,\ldots, J-1.\end{align*}}

@@

Parametric survival models are implemented in a handful of packages:
\pkg{CFC} \citeyearpar{mahani2015bayesian}, \pkg{flexsurv}
\citeyearpar{flexsurv}, \pkg{SmoothHazard} \citeyearpar{smoothHazard},
\pkg{rsptm2} \citeyearpar{clements_liu}, \pkg{mets}
\citeyearpar{scheike2014estimating}, and \pkg{survival}
\citeyearpar{survival-package}. The types of models they allow vary for
each package. For example, \pkg{SmoothHazard} is limited to Weibull
distributions \citeyearpar{smoothHazard}, whereas both \pkg{flexsurv}
and \pkg{survival} allow users to supply any distribution of their
choice. Also, \pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets} and
\pkg{rstpm2} also have the ability to model the effect of time using
splines, which allows flexible modeling of the hazard function.
Moreover, \pkg{flexsurv} has the ability to estimate both scale and
shape parameters for a variety of parametric families. As discussed
above, \pkg{casebase} can model any parametric family whose log-hazard
can be expressed as a linear combination of covariates (including time).
Therefore, our package allows the user to model the effect of time using
splines. Also, by including interaction terms between covariates and
time, it also allows users to fit (non-proportional) time-varying
coefficient models. However, we do not explicitly model any shape
parameter, unlike \pkg{flexsurv}.

Several packages implement penalized estimation for the Cox model:
\pkg{glmnet} \citeyearpar{regpathcox}, \pkg{glmpath}
\citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
\pkg{RiskRegression} \citeyearpar{gerds_blanche}. Moreover, some
packages also include penalized estimation in the context of Cox models
with time-varying coefficients: elastic-net penalization with
\pkg{CoxRidge} \citeyearpar{perperoglou} and \pkg{rstpm2}
\citeyearpar{clements_liu}, while \pkg{survival}
\citeyearpar{survival-package} has an implementation of ridge-penalized
estimation. On the other hand, our package \pkg{casebase} provides
penalized estimation of the hazard function. To our knowledge,
\pkg{casebase} and \pkg{rsptm2} are the only packages to offer this
functionality.

Next, several \proglang{R} packages implement methodologies for
competing risk analysis; for a different perspective on this topic, see
Mahani \& Sharabiani \citeyearpar{mahani2015bayesian}. The package
\pkg{cmprsk} provides methods for cause-specific subdistribution
hazards, such as in the Fine-Gray model
\citeyearpar{fine1999proportional}. On the other hand, the package
\pkg{CFC} estimates cause-specific CIFs from unadjusted, non-parametric
survival functions. Our package \pkg{casebase} also provides
functionalities for competing risk analysis by estimating parametrically
the cause-specific hazards. From these quantities, we can then estimate
the cause-specific CIFs.

Finally, several packages include functions to estimate the CIF. The
corresponding methods generally fall into two categories: transformation
of the estimated hazard function, and semi-parametric estimation of the
baseline hazard. The first category broadly corresponds to parametric
survival models, where the full hazard is explicitly modeled. Using this
estimate, the survival function and the CIF can be obtained using their
functional relationships (see Equations \ref{eqn:surv} and \ref{eqn:CI}
below). Packages providing this functionality include \pkg{CFC},
\pkg{Flexsurv}, \pkg{mets}, and \pkg{survival}. Our package
\pkg{casebase} also follows this approach for both single-event and
competing-risk analyses. The second category outlined above broadly
corresponds to semi-parametric models. These models do not model the
full hazard function, and therefore the baseline hazard needs to be
estimated separately in order to estimate the survival function. This is
achieved using semi-parametric estimators (e.g.~Breslow's estimator) or
parametric estimators (e.g.~spline functions). Packages that implement
this approach include \pkg{RiskRegression}, \pkg{rstpm2}, and
\pkg{survival}. As mentioned in the introduction, a key distinguishing
factor between these two approaches is that the first category leads to
smooth estimates of the cumulative incidence function, whereas the
second category often produces estimates in the form of stepwise
functions. Providing smooth estimates of the CIF was one of the main
motivations for introducing case-base sampling in survival analysis.

@@

Figure \ref{fig:plot-stratified-erspc-data} is built sequentially by
first adding a layer for the area representing the population time in
gray, with subjects having the least amount of observation time plotted
at the top of the y-axis. We immediately notice a distinctive
\emph{stepwise shape} in the population time area. This is due to the
randomization of the Finnish cohorts which were carried out on January 1
of each of year from 1996 to 1999. Coupled with the uniform December 31
2006 censoring date, this led to large numbers of men with exactly 11,
10, 9 or 8 years of follow-up. Tracked backwards in time (i.e.~from
right to left), the population-time plot shows the recruitment pattern
from its beginning in 1991, and the January 1 entries in successive
years. Tracked forwards in time (i.e.~from left to right), the plot for
the first three years shows attrition due entirely to death (mainly from
other causes). Since the Swedish and Belgian centres were the last to
complete recruitment--in December 2003--the minimum potential follow-up
is three years. Tracked further forwards in time (i.e.~after year 3) the
attrition is a combination of deaths and staggered entries. As we can
see, population-time plots summarise a wealth of information about the
study into a simple graph.

@@

\[ S(t) = \exp\left(-\int_0^t \lambda(u;X) du\right),\qquad \lambda(t;X) = \sum_{j=1}^J \lambda_j(t;X).\]
From this, we can derive the event-specific subdensities:

@@

The data contains information on 177 patients who received a stem-cell
transplant for acute leukemia. The event of interest is relapse, but
other competing causes (e.g.~death, progression, graft failure,
graft-versus-host disease) were also recorded. Several covariates were
captured at baseline: sex, disease type (acute lymphoblastic or
myeloblastic leukemia, abbreviated as ALL and AML, respectively),
disease phase at transplant (Relapse, CR1, CR2, CR3), source of stem
cells (bone marrow and peripheral blood, coded as BM+PB, or only
peripheral blood, coded as PB), and age.

@@

We will compare our estimates to that obtained from a corresponding
Fine-Gray model \citeyearpar{fine1999proportional}. The Fine-Gray model
is a semiparametric model for the cause-specific \emph{subdistribution
hazard}, i.e.~the function \(f_k(t)\) such that
\[CI_k(t) =1 - \exp\left( - \int_0^t f_k(u) \textrm{d}u \right),\] where
\(CI_k(t)\) is the cause-specific cumulative incidence. The Fine-Gray
model allows to directly assess the effect of a covariate on the
subdistribution hazard, as opposed to the hazard. For the computation,
we will use the \pkg{timereg} package \citep{timereg}:

@@

\caption{\label{fig:brier1} Comparison of three Cox regression models. (A) Cumulative Incidence curves. (B) Time-dependent Brier scores.}\label{fig:brierpt1}
\end{figure}
\end{CodeChunk}

@@

\caption{\label{fig:cs3lolliPlot} Lollipop plot showing the coefficient estimates between three models. If a coefficient estimate is 0 in any model, it is removed from the plot for simplicity.}\label{fig:coefplots}
\end{figure}
\end{CodeChunk}

These values do not inform us of model performance. To do so, we take
our hazard functions and use them to predict the absolute risk of
everyone in our test set, over time. The probabilities over time for
each individual are averaged, resulting in the absolute risk curves in
Figure \ref{fig:cs3FinalBrier} (A). There is separation between the
curves, but it is minimal. To better demonstrate the differences in
performance between the models, we compare Brier scores over time. In
Figure \ref{fig:cs3FinalBrier} (B), the Brier score is lower in our
penalized logistic regression model compared to both Cox regression
models. However, this difference is minimal. On the other hand, all
three models outperform the Kaplan-Meier model.

@@

\caption{\label{fig:cs3FinalBrier} Four curves comparing the Cumulative Incidence (A) and the Brier scores (B) as they change over time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier.}\label{fig:riskregressionBrier}
\end{figure}
\end{CodeChunk}

@@

As before, we can look at the population-time plot for a graphical
summary of the event incidence. Here, we coloured the exposed
person-time (i.e.~after transplant) in a darker shade of gray. As we can
see, most events occur early during the follow-up period, and therefore
we do not expect the hazard to be constant.

@@

\begin{CodeInput}
R> # Fit several models
R> fit1 <- fitSmoothHazard(fustat ~ exposure,
R+   data = cb_data, time = "futime"
R+ )
R> fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
R+   data = cb_data, time = "futime"
R+ )
R> fit3 <- fitSmoothHazard(fustat ~ exposure * futime,
R+   data = cb_data, time = "futime"
R+ )
\end{CodeInput}
\end{CodeChunk}

Note that the third model includes an interaction term between exposure
and follow-up time. In other words, this model no longer exhibit
proportional hazards. The evidence of non-proportionality of hazards in
the Stanford Heart Transplant data has been widely discussed
\citep{arjas1988graphical}.

@@

\begin{CodeOutput}
#> Model1 Model2 Model3 
#>    827    790    788
\end{CodeOutput}
\end{CodeChunk}

As we can see, the best fit is the third model. By visualizing the
hazard functions for both exposed and unexposed individuals, we can more
clearly see how the hazards are no longer proportional.

@@

\begin{CodeInput}
R> # Compute hazards---
R> # First, create a list of time points for both exposure status
R> hazard_data <- expand.grid(
R+   exposure = c(0, 1),
R+   futime = seq(0, 425,
R+     length.out = 100
R+   )
R+ )
R> 
R> # Set the offset to zero
R> hazard_data$offset <- 0
R> # Use predict to get the fitted values, and exponentiate to
R> # transform to the right scale
R> hazard_data$hazard <- exp(predict(fit3,
R+   newdata = hazard_data,
R+   type = "link"
R+ ))
R> # Add labels for plots
R> hazard_data$Status <- factor(hazard_data$exposure,
R+   labels = c("No Tx", "Tx")
R+ )
R> 
R> ggplot(hazard_data, aes(futime, hazard, colour = Status)) +
R+   geom_line() +
R+   paper_gg_theme +  
R+   ylab("Hazard") +
R+   xlab("Follow-up time (in days)")
\end{CodeInput}

@@

@@

\begin{center}\includegraphics{../figures/unnamed-chunk-9-1} \end{center}

@@

As we can see in the above case-study, the \pkg{casebase} package can
also be used to model time-varying exposures and non-proportional hazard
functions.

@@

This report was generated on 2020-08-05 11:18:44 using the following
computational environment and dependencies:
